---
title: "genotype environment association"
author: "Tram"
date: "7/6/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(raster)
library(data.table)
library(dplyr)
library(stringr)
setwd("/shared/projects/most_kmer/afterkiss/bioclim")
```

## Genotype samples

55 African individuals

```{r}
samples <- fread("./climatic_variables_1.4.csv")
samples <- samples %>% filter(!is.na(Lat))
samples <- samples %>% select(-c(2:21))
samples <- rbind(samples,
                 data.frame(Label = c("C-127", "C-133", "D-7", "D-8"),
                            Long = c(14.53, 14.11, 15.510499, -12.31),
                            Lat = c(1.37, 1.56, 5.485614, 10.35),
                            snp_group = c("C", "A", "C", "D"))
)
write.csv(samples, "./lfmm_samples.csv", quote = F, row.names = F)
```

```{r}
# library(ggmap)
# sbbox <- make_bbox(lon = samples$Long, lat = samples$Lat, f = .1)
# sq_map <- get_map(location = sbbox, maptype = "satellite", source = "google")
# ggmap(sq_map) + 
#   geom_point(data = samples, mapping = aes(x = Long, y = Lat), color = "red3") +
#   geom_density(data = samples, mapping = aes(x = Long, y = Lat))
```

## Climatic data of the African samples

bioclim data from 1970 to 2000 averages (worldclim 2.1)

```{r}
# clim <- getData(name = "worldclim", var = "bio", res = 0.5, lat = na.omit(samples$Lat)[1:2], lon = na.omit(samples$Long)[1:2])
# point <- SpatialPoints(data.frame(long = na.omit(samples$Long)[1:2], lat = na.omit(samples$Lat)[1:2]), proj4string = clim@crs)
# extract(clim, point)

## bioclim data
climfile <- "./wc2.1_30s_bio.zip" # destination path to download the elevation data,  change if needed
download.file("https://geodata.ucdavis.edu/climate/worldclim/2_1/base/wc2.1_30s_bio.zip", 
              destfile = climfile,
              method = "curl")
unzip(climfile)
climdata_dir <- "." # path to the worldclim dataset folder
climdata <- list.files(climdata_dir, pattern = "wc2.1_30s_bio_\\d+.tif", full.names = T)
bioclim <- raster::stack(climdata)

## download elevation data
elevfile <- "./wc2.1_30s_elev.zip" # destination path to download the elevation data,  change if needed
download.file("https://geodata.ucdavis.edu/climate/worldclim/2_1/base/wc2.1_30s_elev.zip", 
              destfile = elevfile,
              method = "curl")
unzip(elevfile)
elev <- raster(gsub(".zip", ".tif", elevfile)) 

## extract values for points
envirvar <- do.call(rbind, lapply(1:nrow(samples), function(i) {
  # clim <- getData(name = "worldclim", var = "bio", res = 0.5, lat = na.omit(samples$Lat)[i], lon = na.omit(samples$Long)[i])
  # elev <- getData(name = "SRTM", lat = na.omit(samples$Lat)[i], lon = na.omit(samples$Long)[i])
  
  point <- SpatialPoints(data.frame(long = na.omit(samples$Long)[i], lat = na.omit(samples$Lat)[i]), proj4string = bioclim@crs)
  clim_values <- raster::extract(bioclim, point)
  elev_value <- raster::extract(elev, point)
  colnames(clim_values) <- gsub("wc2.1_30s_", "", colnames(clim_values))
  df <- data.frame(Label = samples$Label[i], 
                   clim_values, 
                   elevation = elev_value,
                   Long = samples$Long[i], Lat = samples$Lat[i],
                   snp_group = samples$snp_group[i])
  return(df)
}))

write.csv(envirvar[,c(colnames(envirvar)[1], str_sort(colnames(envirvar)[2:20], numeric = T), colnames(envirvar)[21:24])], 
          "./climatic_variables_2.1.csv", row.names = F, append = T)
write.csv(envirvar[,c(colnames(envirvar)[1], str_sort(colnames(envirvar)[2:20], numeric = T), colnames(envirvar)[21])], 
          "lfmm/lfmm_explanatory_2.1.csv",
          row.names = F, append = TRUE)
```

```{r}
library(gplots)
library(viridisLite)
heatmap.2(clim_cor, dendrogram = "row",
          trace = "none", density.info = "none", col = rocket(200))
```



## raster

```{r}
library(raster)
library(rgdal)

af_ctrs <- c("Ghana", "Guinea", "CI", "Togo", "Benin", "Nigeria", 
              "Cameroon", "Gabon", "Congo", "CAF", "COD", "Uganda", "Angola")

all_clim <- getData(name = "worldclim", var = "bio", res = 2.5)
plot(all_clim[[1:3]])

af_geo <- do.call(bind, lapply(af_ctrs, function(ct) {
  print(ct)
  getData("GADM", country = ct, level = 0)
  }))

af_clim <- raster::mask(all_clim, af_geo)
plot(af_clim, xlim = c(-20, 40), ylim = c(-15, 15), axes = F, 
     col = viridis::cividis(100), add = T)

```


```{r}
bioclim <- clim
point <- SpatialPoints(data.frame(long = 13.919876, lat = 1.587391), proj4string = bioclim@crs)
clim_values <- raster::extract(bioclim, point)
```

## try with chelsa database

bioclim data from 1981 to 2010

```{bash}
cd /shared/projects/most_kmer/afterkiss/bioclim
wget --no-host-directories --input-file=envidatS3paths.txt
```

run the codes below with the script `download_bioclim_chelsa_AF.R`
```{r}
climdata_dir <- "./bioclim"
climdata_chelsa <- list.files(climdata_dir, pattern = "CHELSA_bio\\d+_1981-2010_V.2.1.tif", full.names = T)
bioclim_chelsa <- raster::stack(climdata_chelsa)

envirvar_chelsa <- do.call(rbind, lapply(1:nrow(samples), function(i) {
  # clim <- getData(name = "worldclim", var = "bio", res = 0.5, lat = na.omit(samples$Lat)[i], lon = na.omit(samples$Long)[i])
  # elev <- getData(name = "SRTM", lat = na.omit(samples$Lat)[i], lon = na.omit(samples$Long)[i])
  print(i)
  point <- SpatialPoints(data.frame(long = na.omit(samples$Long)[i], lat = na.omit(samples$Lat)[i]), proj4string = bioclim_chelsa@crs)
  clim_values <- raster::extract(bioclim_chelsa, point)
  elev_value <- raster::extract(elev, point)
  colnames(clim_values) <- gsub("(CHELSA_|_1981.2010_V.2.1)", "", colnames(clim_values))
  df <- data.frame(Label = samples$Label[i], 
                   clim_values, 
                   elevation = elev_value,
                   Long = samples$Long[i], Lat = samples$Lat[i],
                   snp_group = samples$snp_group[i])
  return(df)
}))
```

compare chelsa and the worldclim data 
```{r}
envirvar_chelsa <- read.csv("climatic_variables_af_chelsa.csv")
write.csv(envirvar_chelsa[,1:21], 
          "/shared/projects/most_kmer/afterkiss/gea/lfmm/lfmm_explanatory_af_chelsa.csv",
          quote = F, row.names = F)
for (i in c(2:21)) print(cor(envirvar[,i], envirvar_chelsa[,i]))
```

## Climatic data of the occurences in Vietnam

```{r}
occ_file <- "occ_coffee_vnm.csv"
occurence <- read.csv(occ_file)
```
get bioclim variables for the occurence with the script `download_bioclim_chelsa_VN.R` 


## future data

- download data in 2050 (CMIP5), model BCC-CSM2-MR, pathway ssp585, resolution 30s

```{bash}
wget https://geodata.ucdavis.edu/cmip6/5m/BCC-CSM2-MR/ssp585/wc2.1_5m_bioc_BCC-CSM2-MR_ssp585_2041-2060.tif
```

```{r}
download.file("https://biogeo.ucdavis.edu/data/climate/cmip5/30s/bc85bi50.zip", 
              destfile = "bc85bi50.zip",
              method = "curl")
```
FAILED

- download data in period 2041-2060 (CMIP6), model BCC-CSM2-MR, pathway ssp585, resolution 30s

```{bash}
wget https://geodata.ucdavis.edu/cmip6/5m/MIROC6/ssp585/wc2.1_5m_bioc_MIROC6_ssp585_2041-2060.tif
```

```{r}
library(geodata)
cmip6_tile(lon = occurence$Lon[1],
           lat = occurence$Lat[1],
           model = "BCC-CSM2-MR",
           ssp = "585",
           time = "2041-2060",
           var = "bioc",
           res = 2.5,
           path = tempdir())

raster::getData('CMIP5', var='tmin', res=10, rcp=85, model='AC', year=70) 
```

FAILED

- download data in period 2041-2070 (CMIP6), model IPSL-CM6A-LR, pathway ssp585, resolution 30s

from CHELSA
```{bash}
cd /shared/projects/most_kmer/afterkiss/bioclim
wget --no-host-directories --input-file=envidatS3paths_future_ipsl.txt
```

from worldclim
```{bash}
wget --no-check-certificate https://geodata.ucdavis.edu/cmip6/30s/IPSL-CM6A-LR/ssp585/wc2.1_30s_bioc_IPSL-CM6A-LR_ssp585_2041-2060.tif
```


- download data in period 2041-2070 (CMIP6), model GFDL-ESM4, pathway ssp585, resolution 30s

```{bash}
wget --no-host-directories --input-file=envidatS3paths_future_gfdl.txt
```

