---
title: "Detection of k-mers associated to bioclim variables"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(data.table)
library(tidyverse)
library(Biostrings)
library(karyoploteR)
```


This work is mainly done on IFB cluster.


## Objectives
The purpose is to detect kmers associated with bioclim variables (and maybe also elevation data), using the native samples in Africa.    
Firstly, the Kiss pipeline will be applied to obtain a kmer data set, and detect kmers under selection (associated with population structure, that will be useful for the association test).   
Secondly, I will apply lfmm with different strategies (the whole kmer data set, the putatively adaptive kmers, putatively adaptive kmers + neutral kmers). Then, I will check the distribution of the p-values and/or q-values to determine which strategy is better in terms of minimizing false positives. 
Finally, the significant kmers will be mapped on the reference genome and assembled for annotations ... (to be developed)


## Previously sequenced samples

55 African individuals were sequenced using Illumina.  
They are representative of the 8 genetic groups and consist of some admixed individuals.  

They were cleaned by filtering adapters and trimming low-quality read bases.

removing adapters by fastp 0.23.1: 
`fastp --detect_adapter_for_pe --disable_quality_filtering --disable_length_filtering`  
trimming read bases < Q20 (min read length = 35) by cutadapt 3.1: 
`cutadapt -m 35 -q 20,20`

The clean reads are available in `/data/projects/rob/data/raw_fastq_2019` on iTrop,
and `/shared/projects/elai_most/data/fastq/fastq_2019_clean` on IFB.

```{r}
sum_fastqc <- function(stat_file) {
  all_stats <- read.delim(stat_file)
  
  min_read_length <- all_stats %>% 
    pull(FastQC_mqc.generalstats.fastqc.avg_sequence_length) %>% 
    min()
  max_read_length <- all_stats %>% 
    pull(FastQC_mqc.generalstats.fastqc.avg_sequence_length) %>% 
    max()
  
  min_num_seq <- all_stats %>% 
    pull(FastQC_mqc.generalstats.fastqc.total_sequences) %>% 
    min()
  max_num_seq <- all_stats %>% 
    pull(FastQC_mqc.generalstats.fastqc.total_sequences) %>% 
    max()
  
  min_cov <- all_stats %>% 
    mutate(coverage = FastQC_mqc.generalstats.fastqc.total_sequences*2*FastQC_mqc.generalstats.fastqc.avg_sequence_length/750000000) %>% 
    pull(coverage) %>% min
  max_cov <- all_stats %>% 
    mutate(coverage = FastQC_mqc.generalstats.fastqc.total_sequences*2*FastQC_mqc.generalstats.fastqc.avg_sequence_length/750000000) %>% 
    pull(coverage) %>% max
  
  cat("- Read length:", min_read_length, "-", max_read_length, "\n")
  cat("- Number of sequences:", min_num_seq*2, "-", max_num_seq*2, "\n")
  cat("- Coverage:", min_cov, "-", max_cov, "\n")
}
```


```{r}
sum_fastqc("/shared/projects/afrob_seq/wgs_snp/vcf_bgi_2021/gbp_output/fastqc/multiqc/fastqc_data/multiqc_general_stats.txt")
```


Clean reads    
- Read length: 88 - 150
- Number of sequences: 76M - 704M   
- Coverage: 10 - 92    
- Base quality: > 25    
- No adapter contamination    

Make a soft link to a seperate folder to run Kiss more easily

```{bash}
kiss_fastq=/shared/projects/most_kmer/kiss_af/fastq
mkdir -p $kiss_fastq
ln -s /shared/projects/elai_most/data/fastq/fastq_2019_clean/[^TR,^S]*/*.f*q.gz $kiss_fastq/
```


## Newly sequenced samples

We selected 5 African samples with more pure genetic structure based on snmf results using kasp SNPs (together with other 15 African + 2 VN samples).    
These samples were degraded (level B or C on BGI quality report, and so some digestion step was skipped).   
Therefore, we need to check carefully the sequecing quality.    
The data is currently stored in `/shared/projects/afrob_seq/fastq_2022` on IFB

```{bash rename_fastq}
fastqfiles=$(ls /shared/projects/afrob_seq/fastq_2022/Reads/Clean/*/*)
newpath=/shared/projects/afrob_seq/fastq_2022/renamed
mkdir $newpath
for f in $fastqfiles
do
    sample=$(echo $(basename $(dirname $f)) | sed 's/A//')
    read=$(echo ${f:(-7)})
    mkdir -p $newpath/$sample
    ln -s $f $newpath/$sample/$sample"_"$read
done
ll $newpath
```

```{bash quality_control}
mkdir -p /shared/projects/afrob_seq/fastq_2022/quality
cd /shared/projects/afrob_seq/fastq_2022/quality
nano qc_config.yaml
nano run_qc.sh
sbatch run_qc.sh
```

```{r}
sum_fastqc("/shared/projects/afrob_seq/fastq_2022/quality/fastqc/fastqc_data/multiqc_general_stats.txt")
```


The reads were preprocessed and cleaned by BGI (22 samples)    
- Read length: 150
- Number of read bases: 44M - 248M   
- Coverage: 9 - 49    
- Base quality: > 25    
- No adapter contamination    



Even though the reads are of good quality, I keep cleaning them in the same way as for the previous data.   
The scripts for cleaning the previous data is [here](/shared/projects/afrob_seq/wgs_snp/vcf_bgi_2021/config/clean_fastq_2019.sh)    


```{bash}
cd ..
mkdir cleaned_reads
cd cleaned_reads
nano clean_fastq.sh
sbatch clean_fastq.sh
```


Check the quality of the cleaned reads

```{bash}
cd ../
mkdir quality_cleaned
cp quality/qc_config.yaml quality_cleaned/
cp quality/run_qc.sh quality_cleaned/
nano quality_cleaned/qc_config.yaml 
nano quality_cleaned/run_qc.sh
sbatch quality_cleaned/run_qc.sh
```

Cleaned reads:
```{r}
sum_fastqc("/shared/projects/afrob_seq/fastq_2022/quality_cleaned/fastqc/fastqc_data/multiqc_general_stats.txt")
```

- Read length: 148.2436 - 149.3562 
- Number of sequences: 44M - 247M
- Coverage: 8.827241 - 49.28375

Make a soft link of 5 new samples to the fastq folder for kiss

```{bash}
kiss_fastq=/shared/projects/most_kmer/kiss_af/fastq
mkdir -p $kiss_fastq
ln -s /shared/projects/afrob_seq/fastq_2022/renamed/[C,D]*/*.f*q.gz $kiss_fastq/
fastq_files=$(ls $kiss_fastq)

# rename the new seqs to have homogenous name format for all the samples
for f in $fastq_files 
do
    if [[ $f =~ fq ]]
    then 
        echo -e "$f"
        sample=$(echo $f | cut -d "_" -f 1)
        read=$(echo $f | cut -d "_" -f 2 | cut -d "." -f 1)
        new_name=$sample"_R"$read".fastq.gz"
        echo -e $new_name
        cd $kiss_fastq
        mv $f $new_name
    fi
done
```


## Climatic data

Climatic data for the African samples was extracted by `climatic_data.Rmd` and stored in `gea/climatic_variables_2.1.csv`


## Extract k-mers and detect k-mers under selection

results in `/shared/projects/most_kmer/kiss_af`  


1,856,279,562 k-mers in total  
410,304,144 k-mers under selection


## LFMM results using only outlier kmers

check if all the runs are done
```{r}
lfmm_out <- "./gea/lfmm/out_K5_outlier"
real <- dirname(list.files(lfmm_out, "candidates_kmers.csv", full.names = F, recursive = T))
length(real)

n <- 1856
expect <- sapply(1:n, function(r) {
  if (r %% 100 == 0) {
    b <- r/100 - 1
    s <- 100
  } else {
    b <- r %/% 100
    s <- r %% 100
  }
  return(paste(b, s, sep = "_"))
})
length(expect)

# files that haven't finished
expect[!expect %in% real]
```

gather all the lfmm results for all the k-mer batches using the log files (faster i think)

```{r}
log_K5 <- "./gea/lfmm/log_K5_outlier"
log_files <- list.files(log_K5, pattern = ".log", full.names = T)
n <- length(log_files)

cand_kmers <- do.call(rbind, lapply(1:n, function(i) {
  cat(i, "\n")
  df <- fread(log_files[i], skip = 12, nrows = 21, sep = "\t", 
              data.table = FALSE)
  df <- df %>% mutate(batch = i)
  return(df)
  })
  )

sum_cand_kmers <- cand_kmers %>% 
  dplyr::group_by(variable) %>% 
  summarise(across(significant_kmers:significant_outliers, ~ sum(.x))) %>% 
  arrange(desc(significant_outliers))

format(as.data.frame(sum_cand_kmers), big.mark = ",")

can_files <- list.files("./gea/lfmm/out_K5_outlier", 
                        pattern = "candidates_kmers.csv", 
                        full.names = T, recursive = T)
can_count <- lapply(can_files, fread)
all_can <- do.call(rbind, can_count)
dim(all_can)
```


    variable significant_kmers significant_outliers
1      bio_2        30,012,712           30,012,712
2      bio_7        27,404,828           27,404,828
3      bio_3         9,000,933            9,000,933
4     bio_15         8,688,835            8,688,835
5     bio_14         2,803,076            2,803,076
6      bio_9         2,052,938            2,052,938
7      bio_1         1,942,297            1,942,297
8      bio_6         1,874,238            1,874,238
9     bio_17         1,844,878            1,844,878
10     bio_5         1,672,097            1,672,097
11    bio_12         1,498,292            1,498,292
12    bio_13         1,435,270            1,435,270
13    bio_11         1,343,750            1,343,750
14     bio_8         1,150,733            1,150,733
15 elevation           909,767              909,767
16     bio_4           645,615              645,615
17    bio_10           508,358              508,358
18    bio_18           355,445              355,445
19    bio_16           174,929              174,929
20    bio_19            34,923               34,923
in total: 48,995,113

```{r}
log_K5_chelsa <- "./gea/lfmm/log_K5_outlier_chelsa/"
log_K5_chelsa_files <- list.files(log_K5_chelsa, pattern = ".log", full.names = T)
n <- length(log_K5_chelsa_files)

cand_kmers_chelsa <- do.call(rbind, lapply(1:n, function(i) {
  cat(i, "\n")
  df <- fread(log_K5_chelsa_files[i], skip = 12, nrows = 21, sep = "\t", 
              data.table = FALSE)
  df <- df %>% mutate(batch = i)
  return(df)
  })
  )

sum_cand_kmers_chelsa <- cand_kmers_chelsa %>% 
  dplyr::group_by(variable) %>% 
  dplyr::summarise(across(significant_kmers:significant_outliers, ~ sum(.x))) %>% 
  arrange(desc(significant_outliers))

format(as.data.frame(sum_cand_kmers_chelsa), big.mark = ",")

can_files_chelsa <- list.files("./gea/lfmm/out_K5_outlier_chelsa/", 
                        pattern = "candidates_kmers.csv", 
                        full.names = T, recursive = T)
can_count_chelsa <- lapply(can_files_chelsa, fread)
all_can_chelsa <- do.call(rbind, can_count_chelsa)
# number of significant kmers
nrow(all_can_chelsa)
# number of significant outliers
length(all_can_chelsa$sequence[all_can_chelsa$outlier])
```


## LFMM results using all kmers

check if all the runs are done
```{r}
lfmm_out <- "./gea/lfmm/out_K5"
real <- dirname(list.files(lfmm_out, "candidates_kmers.csv", full.names = F, recursive = T))
length(real)

n <- 1856
expect <- sapply(1:n, function(r) {
  if (r %% 100 == 0) {
    b <- r/100 - 1
    s <- 100
  } else {
    b <- r %/% 100
    s <- r %% 100
  }
  return(paste(b, s, sep = "_"))
})
length(expect)

# files that haven't finished
expect[!expect %in% real]
```

gather all the lfmm results for all the k-mer batches using the log files (faster i think)

```{r}
log_K5 <- "./gea/lfmm/log_K5"
log_files <- list.files(log_K5, pattern = ".log", full.names = T)
n <- length(log_files)

cand_kmers <- do.call(rbind, lapply(1:n, function(i) {
  cat(i, "\n")
  df <- fread(log_files[i], skip = 12, nrows = 21, sep = "\t", 
              data.table = FALSE)
  df <- df %>% mutate(batch = i)
  return(df)
  })
  )

sum_cand_kmers <- cand_kmers %>% 
  dplyr::group_by(variable) %>% 
  summarise(across(significant_kmers:significant_outliers, ~ sum(.x))) %>% 
  arrange(desc(significant_outliers))

format(as.data.frame(sum_cand_kmers), big.mark = ",")

can_files <- list.files("./gea/lfmm/out_K5", 
                        pattern = "candidates_kmers.csv", 
                        full.names = T, recursive = T)
can_count <- lapply(can_files, fread)
all_can <- do.call(rbind, can_count)
# number of significant kmers
nrow(all_can)
# number of significant outliers
length(all_can$sequence[all_can$outlier])
```


    variable significant_kmers significant_outliers
1      bio_7        46,295,609           15,209,945
2      bio_2        39,549,887           12,989,670
3      bio_6        24,400,908            8,025,601
4      bio_5        22,509,920            7,403,012
5      bio_3        21,516,670            7,272,921
6     bio_15        21,693,587            7,066,614
7     bio_18         8,244,191            2,901,733
8     bio_12         7,338,038            2,653,891
9     bio_11         6,753,399            1,923,549
10     bio_8         4,433,745            1,534,612
11     bio_1         3,317,872            1,147,873
12    bio_13         2,374,762              957,389
13     bio_9         2,805,126              907,330
14    bio_17         2,199,302              733,119
15    bio_14         1,270,101              501,134
16 elevation         1,394,007              479,655
17     bio_4           981,808              222,915
18    bio_10           107,796               38,276
19    bio_19            36,104                3,483
20    bio_16                 0                    0
105,180,163

-> use significant kmers from output of lfmm on all kmers, because this takes into account both outliers and neutral kmers

## Assembly of associated k-mers using mergeTags

try [mergeTags](https://github.com/Transipedia/dekupl-mergeTags)

install mergeTags
```{bash}
ssh baotram@core.cluster.france-bioinformatique.fr
cd /shared/ifbstor1/projects/most_kmer/afterkiss/gea
mkdir assembly
cd assembly
git clone https://github.com/Transipedia/dekupl-mergeTags.git
cd dekupl-mergeTags
make
```

test mergeTags
```{bash}
mkdir test
cd test
nano merged_lfmm_pvalues.csv
awk '{print $1"\t"$4"\t"$4"\t"$4"\t"$3}' merged_lfmm_pvalues.csv > merged_lfmm_pvalues2mergetags.csv
../dekupl-mergeTags/mergeTags -k 31 -m 15 -n merged_lfmm_pvalues2mergetags.csv > out_merged_lfmm_pvalues2mergetags.csv
```

run mergeTags on the all associated k-mers (with minimum overlap = 15)


```{r}
fwrite(all_can %>% relocate(sequence), 
       "./assembly/significant_kmers_lfmm.tsv", 
       sep = "\t", row.names = F)
```




```{bash}
sbatch /shared/ifbstor1/projects/most_kmer/afterkiss/gea/assembly/mergeTags_assembly.sh
```

### with m 15 (default)
```{r}
mergeTags_out <- "./assembly/mergeTags_significant_kmers_lfmm_m15.tsv"
mT_contigs <- fread(mergeTags_out, data.table = F)
unassembled <- mT_contigs %>% 
  filter(nb_merged_kmers == 1)
contigs <- mT_contigs %>% 
  filter(nb_merged_kmers != 1)
```
-> 14,842,939 unassembled k-mers
-> 21,841,084 contigs


analyse contigs
```{r}
contigs <- contigs %>% 
  # filter(nb_merged_kmers > 100) %>% 
  mutate(contig_length = str_count(contig))

sum_merged_kmers <- contigs %>% 
  mutate(nb_kmers_ranges = cut(nb_merged_kmers,
                               c(min(contigs$nb_merged_kmers), 
                                 seq(100, 500, 100), 
                                 max(contigs$nb_merged_kmers)),
                               dig.lab = 5,
                               include.lowest = T, right = F)) %>% 
  group_by(nb_kmers_ranges) %>% 
  summarise(nb_kmers = n())

sum_ct_length <- contigs %>% 
  mutate(contig_length_ranges = cut(contig_length, 
                                    c(min(contigs$contig_length), 500,
                                      1000, 1500, 2000,
                                      max(contigs$contig_length)),
                                    dig.lab = 5,
                                    include.lowest = T, right = F)) %>% 
  group_by(contig_length_ranges) %>% 
  summarise(nb_contigs = n())
```


```{r}
# number of contigs associated with each bioclim variables
bio_contigs <- contigs %>% 
  select(bio_1:elevation) %>% 
  summarise(across(everything(), ~ sum(.)))
contigs %>% 
  rowwise() %>% 
  mutate(n_bio = across(bio_1:elevation, ~ sum(.)))
```

```{r}
long_contigs <- contigs %>% 
  filter(contig_length > 100)
long_contigs$contig_id <- paste0("contig_", 1:nrow(long_contigs))
```




```{r}
hist(contigs$nb_merged_kmers, breaks = 1000)
ggplot(contigs %>% filter(nb_merged_kmers > 100)) +
  geom_histogram(aes(x = nb_merged_kmers), binwidth = 100) +
  theme_minimal() +
  scale_y_sqrt(breaks = c(0, 10, 100, 1000, 2000, 3000, 4000)) 
min(contigs$nb_merged_kmers)
max(contigs$nb_merged_kmers)


ggplot(contigs) +
  geom_histogram(aes(x = contig_length), binwidth = 100) +
  theme_minimal() +
  scale_y_sqrt() 
min(contigs$contig_length)
max(contigs$contig_length)
```


```{r}
long_contigs <- contigs %>% 
  filter(contig_length > 60)

ggplot(long_contigs) +
  geom_histogram(aes(x = nb_merged_kmers), binwidth = 100) +
  theme_minimal() +
  scale_y_sqrt(breaks = c(0, 10, 100, 1000, 5000, 10000)) 

ggplot(long_contigs) +
  geom_histogram(aes(x = contig_length), binwidth = 100) +
  theme_minimal() +
  scale_y_sqrt()
```


### with m 21

```{r}
mergeTags21_out <- "./assembly/mergeTags_significant_kmers_lfmm_m21.tsv"
mT_res_21 <- fread(mergeTags21_out, data.table = F)
# unassembled_m30 <- mT_res_30 %>% 
#   filter(nb_merged_kmers == 1)
contigs_m21 <- mT_res_21 %>% 
  filter(nb_merged_kmers != 1) %>% 
  # filter(nb_merged_kmers > 61) %>% 
  mutate(contig_length = str_count(contig))
nrow(mT_res_21) - nrow(contigs_m21)
```
-> 30,645,634 unassembled kmers
-> 22,767,142 contigs



```{r}
long_contigs_m21 <- contigs_m21 %>% 
  filter(contig_length > 60)
nrow(long_contigs_m21)
```
-> 964,556 contigs >= 61 bp

```{r}
ggplot(contigs_m21) +
  geom_histogram(aes(x = nb_merged_kmers), binwidth = 100) +
  theme_minimal() +
  scale_y_sqrt(breaks = c(0, 10, 100, 1000, 2000, 3000, 4000)) 

min(contigs_m21$nb_merged_kmers)
max(contigs_m21$nb_merged_kmers)


ggplot(contigs_m21) +
  geom_histogram(aes(x = contig_length), binwidth = 100) +
  theme_minimal() +
  scale_y_sqrt() 
```


```{r}
contigs_m21 %>% 
  filter(nb_merged_kmers > 30)

sum_merged_kmers_m21 <- contigs_m21 %>% 
  mutate(nb_kmers_ranges = cut(nb_merged_kmers,
                               c(min(contigs_m21$nb_merged_kmers), 
                                 seq(10, 150, 10), 
                                 max(contigs_m21$nb_merged_kmers)),
                               dig.lab = 5,
                               include.lowest = T, right = F)) %>% 
  group_by(nb_kmers_ranges) %>% 
  summarise(nb_contigs = n())

sum_ct_length_m21 <- contigs_m21 %>% 
  mutate(contig_length_ranges = cut(contig_length, 
                                    c(min(contigs_m21$contig_length),
                                      seq(50, 500, 50), 
                                      max(contigs_m21$contig_length)),
                                    dig.lab = 5,
                                    include.lowest = T, right = F)) %>% 
  group_by(contig_length_ranges) %>% 
  summarise(nb_contigs = n())
```

```{r}
ggplot(sum_merged_kmers_m21, aes(x = nb_kmers_ranges, y = nb_contigs)) + 
  geom_col() + 
  geom_text(aes(label = nb_contigs), vjust = -1) + 
  scale_y_sqrt() + 
  theme_minimal() +
  xlab("Number of merged k-mers in contigs") +
  ylab("Number of contigs")

ggplot(sum_ct_length_m21, aes(x = contig_length_ranges, y = nb_contigs)) + 
  geom_col() + 
  geom_text(aes(label = nb_contigs), vjust = -1) + 
  scale_y_sqrt() + 
  theme_minimal() +
  xlab("Contig length") +
  ylab("Number of contigs")
```

## Assembly of associated k-mers using abyss

```{r}
kmer_set <- BStringSet(all_can$sequence)
names(kmer_set) <- paste0("kmer_", 1:length(kmer_set))
writeXStringSet(kmer_set, "./assembly/significant_kmers.fasta")
```

## Blast of assembled contigs

## for contigs assembled by m15
```{r}
long_contigs$contig_id <- paste0("contig_", 1:nrow(long_contigs))
contigs4blast <- BStringSet(long_contigs$contig)
names(contigs4blast) <- long_contigs$contig_id

writeXStringSet(contigs4blast, "./assembly/contigs_m15.fasta", format = "fasta")
```


```{r}
blast_out <- "./assembly/all_contigs_m15_blast_result.txt"
blast_contigs <- fread(blast_out)
colnames(blast_contigs) <- c("contig_id", "mapped_chr", "start", "end", "mapped_start", "mapped_end", "identity", "gaps", "mismatches", "align_length")

blast_contigs %>% 
  # filter(evalue < 0.01) %>% 
  arrange(desc(align_length))

long_contigs %>% 
  filter(contig_id %in% blast_contigs$contig_id) %>% 
  slice_max(contig_length)
long_contigs %>% 
  filter(contig_id == "contig_611870")

blast_contigs %>% 
  filter(contig_id == "contig_611870") %>% 
  group_by(mapped_chr) %>% 
  summarise(n = n())

blast_contigs %>% 
  filter(!grepl("Contig", mapped_chr)) %>% 
  group_by(contig_id) %>% 
  summarise(n_hits = n()) %>% 
  arrange(n_hits) %>% 
  filter(n_hits == 1)

blast_contigs %>% 
  filter(contig_id == "contig_3531067") %>% 
  # filter(contig_id %in% blast_contigs$contig_id) %>% 
  group_by(mapped_chr) %>% 
  summarise(n = n())

blast_contigs %>% 
  filter(align_length > 100)
```


## for contigs assembled by m21
```{r}
long_contigs_m21$contig_id <- paste0("contig_", 1:nrow(long_contigs_m21))
contigs4blast <- BStringSet(long_contigs_m21$contig)
names(contigs4blast) <- long_contigs_m21$contig_id

writeXStringSet(contigs4blast, "./assembly/contigs_m21.fasta", format = "fasta")
```

```{r}

```

